---
title: "p8105_hw2_Coo2124"
output: github_document
---

```{r setup, include=FALSE}
library(tidyr)
library(dplyr)
```

#Problem1 
```{r}
data1 <- read.csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv", header = T)
transit_new <- data1 %>%
  select(Line, Station.Name, Station.Latitude, Station.Longitude, Route1:Route11, Entry, Vending, Entrance.Type, ADA) %>%
  mutate(across(starts_with("Route8"):starts_with("Route11"), as.character),
         Entry = ifelse(Entry == "YES", TRUE, FALSE)) %>%
  pivot_longer(cols = starts_with("Route"),
               names_to = "Route_Number",   
               values_to = "Route") %>%
  distinct()
View(transit_new)
dim(transit_new)
str(transit_new)

```

The NYC Transit dataset contains `r nrow(data)` rows and `r ncol(data)` columns, giving us detailed info about subway entrances and exits. Key variables include `line`, `station_name`, `latitude`, `longitude`, `routes_served`, `entry`, `vending`, `entrance_type`, and `ADA_compliance`. I converted entry from a character format `("YES"/"NO")` to a logical format `(TRUE/FALSE)` using `ifelse` in R. For cleaning, I kept only the relevant columns mentioned above. The data looks tidy—each row is one observation of an entrance or exit, with each variable neatly in its own column.


#1. How many distinct stations are there?
```{r}
distinct_stations <- transit_new %>%
  distinct(Line, Station.Name) %>%
  count()
print(distinct_stations)
```
#answer: There are 465 distinct stations

#2.How many stations are ADA compliant?
```{r}
ada_compliant_stations_count <- transit_new %>%
  filter(ADA == TRUE) %>%                         
  distinct(Line, Station.Name) %>%                 
  nrow()                                           

print(ada_compliant_stations_count)
```
#answer:There are 84 station in ADA compliant

#3.What proportion of station entrances / exits without vending allow entrance?
```{r}
proportion_no_vending_entry <- transit_new %>%
  filter(Vending == "NO") %>%                      
  summarise(Proportion = mean(Entry == TRUE, na.rm = TRUE)) 

print(proportion_no_vending_entry)
```
#answer: The proportion of station entrances/exits without vending that allow entrance is 38%.

#Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train?
```{r}
data1[, paste0("Route", 8:11)] <- lapply(data1[, paste0("Route", 8:11)], as.character)

distinct_a_stations <- data1 %>%
  pivot_longer(
    cols = Route1:Route11,             
    names_to = "route_num",           
    values_to = "route"              
  ) %>%
  filter(route == "A") %>%
  select(Station.Name, Line) %>%  
  distinct()

nrow(distinct_a_stations)
```

#answer: There are 60 distinct stations that serve the A train in the NYC subway system.

#Of the stations that serve the A train, how many are ADA compliant?
```{r}
distinct_a_stations_ada <- data1 %>%
  pivot_longer(
    cols = Route1:Route11,
    names_to = "route_num",
    values_to = "route"
  ) %>%
  filter(route == "A", ADA == TRUE) %>%  # Filter for ADA compliance
  select(Station.Name, Line) %>%
  distinct()

nrow(distinct_a_stations_ada)
```

#answer:Of the stations that serve the A train, 17 stations are ADA compliant.



